For Induction-Induction, Induction is Enough

### 2019-11-11 paper/paper.tex
stuff

Inductive-inductive types (IITs \cite{forsberg-phd}) allow the mutual
definition of a type and a family of types over the first one. IITs
were originally introduced to represent the well-typed syntax of type
theory itself, and the main example is still the Chapman-style
\cite{chapman09eatitself} syntax for a type theory. A minimal use case
is the IIT of contexts and types given by the following constructors.
\begin{alignat*}{5}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\Set \\
  & \mathsf{empty} && : \Con \\
  & \mathsf{ext} && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Gamma \\
  & \mathsf{El} && : (\Gamma:\Con)\ra\Ty\,(\mathsf{ext}\,\Gamma\,(\mathsf{U}\,\Gamma))
%  & \mathsf{Pi} && : (\Gamma:\Con)(A:\Ty\,\Gamma)\ra\Ty\,(\mathsf{ext}\,\Gamma\,A)\ra\,\Ty\,\Gamma
\end{alignat*}
This type has two sorts, $\Con$ and $\Ty$. The $\mathsf{ext}$
constructor of $\Con$ refers to $\Ty$ and the $\Ty$-constructor
$\mathsf{U}$ refers to $\Con$, hence the two sorts have to be defined
at the same time. Moreover, $\Ty$ is indexed over $\Con$, which
precludes the above reduction which worked for
$\mathsf{isEven}$--$\mathsf{isOdd}$, as we would get a type indexed
over itself. Another unique feature of IITs (which also shows up in
higher inductive types \cite{HoTTbook}) is that later constructors can
refer to previous constructors: in our case, $\mathsf{El}$ mentions
$\mathsf{ext}$.
% \[
%   \mathsf{Con?} : (b:\mathsf{Bool})\ra\mathsf{if}\,b\,\mathsf{then}\,\Set\,\mathsf{else}\,(\mathsf{Con?}\,\mathsf{false}\ra\Set),
% \]

@@

As opposed to mutual inductive types, reducing IITs to inductive types is an
open problem. Forsberg \cite{forsberg-phd} showed a reduction which only
supports a simpler, non-recursive-recursive elimination principle
(\todo{check}). Hugunin \cite{jasper} reduced several IITs to inductive types,
working inside a UIP-free cubical type theory, but he also only constructed the
simpler non-recursive-recursive eliminator.

The theory of signatures in \cite{Kaposi:2019:CQI:3302515.3290315} is
a small type theory in which every context encodes a valid quotient
inductive-inductive type (QIIT) signature. For example, the signature
for natural numbers is simply the context
$(Nat:\U,zero:Nat,suc:Nat\ra Nat)$ of length three ($Nat$, $zero$ and
$suc$ are variable names). \cite{Kaposi:2019:CQI:3302515.3290315}
showed that if this theory of signatures exists, then any finitary
QIIT can be defined from it. If we omit the type former $\mathsf{Id}$
from the theory of signatures, we get a theory of finitary IIT
signatures, and the construction of
\cite{Kaposi:2019:CQI:3302515.3290315} still works: constructing an
arbitrary IIT from the theory of IIT signatures does not need the
$\mathsf{Id}$ type former to be present.

In this paper, we construct the theory of IIT signatures only using
indexed inductive types, thus \todo{rewrite from here}

To derive the existence of all IITs from the theory of IIT signatures
we use a variant of the CwF model
\cite{Kaposi:2019:CQI:3302515.3290315} where types are modelled not by
displayed CwFs but by CwF isofibrations. We need this extension to
support infinitary constructors. Furthermore, using the method
introduced by Streicher \cite{streichersemantics} we show the
existence of the theory of IIT signatures. As opposed to Streicher, we
don't use quotients in this construction as the theory of IIT
signatures can be given only using neutral terms.

Our metatheory is ETT. Using Hofmann \cite{}, Theo \cite{} translation
we can translate our arguments to ITT+UIP+funext. Section
\ref{sec:ambroise} was formalised in Agda using UIP, rewrite rules and
funext.

This is the paper version of our TYPES 2019 talk.

@@

We first define a \emph{theory of IIT signatures} in which every context is a
valid IIT signature (Section \ref{sec:theory_of_signatures}). We present this
type theory as an algebraic \cite{ttintt} theory, specifically as a quotient
inductive-inductive type (QIIT). In Section \ref{sec:ambroise} we show that this
QIIT can be constructed only using inductive types. In Section
\ref{sec:constructingiits} we describe the construction of IITs from the syntax
of IIT signatures.

@@

The theory of IIT signatures is given as an initial category with families
(CwF)\cite{Dybjer96internaltype} with certain additional type formers. The CwF
part constitutes the substitution calculus, while the type formers allow
declaring new sorts and operators in a signature.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \Con && : \Set \\
  & \Ty  && : \Con\ra\Set \\
  & \Sub  && : \Con\ra\Con\ra\Set \\
  & \Tm  && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set \\
  & \id && : \Sub\,\Gamma\,\Gamma \\
  & \blank\circ\blank && : \Sub\,\Theta\,\Delta\ra\Sub\,\Gamma\,\Theta\ra\Sub\,\Gamma\,\Delta \\
  & \ass && : (\sigma \circ \delta) \circ \nu = \sigma \circ (\delta \circ \nu) \\
  & \idl && : \id\circ\sigma = \sigma \\
  & \idr && : \sigma\circ\id = \sigma \\
  & \blank[\blank] && : \Ty\,\Delta\ra\Sub\,\Gamma\,\Delta\ra\Ty\,\Gamma \\
  & \blank[\blank] && : \Tm\,\Delta\,A\ra(\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma]) \\
  & [\id] && : A[\id] = A \\
  & [\circ] && : A[\sigma \circ \delta] = A[\sigma][\delta] \\
  & [\id] && : t[\id] = t \\
  & [\circ] && : t[\sigma \circ \delta] = t[\sigma][\delta] \\
  & \boldsymbol{\cdot} && : \Con \\
  & \epsilon && : \Sub\,\Gamma\,\boldsymbol{\cdot} \\
  & {\boldsymbol{\cdot}\eta} && : \{\sigma : \Sub\,\Gamma\,\boldsymbol{\cdot}\} \ra \sigma = \epsilon \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \blank,\blank && : (\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma])\ra\Sub\,\Gamma\,(\Delta\ext A) \\
  & \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
  & \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
  & {\ext\beta_1} && : \p\circ(\sigma, t) = \sigma \\
  & {\ext\beta_2} && : \q[\sigma, t] = t \\
  & {\ext\eta} && : (\p, \q) = \id \\
  & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\[0.5em]
  & \rlap{(2) Universe} \\[0.5em]
  & \U && : \Ty\,\Gamma \\
  & \El && : \Tm\,\Gamma\,\U\ra\Ty\,\Gamma \\
  & {\U[]} && : \U[\sigma] = \U \\
  & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\[0.5em]
  & \rlap{(3) Inductive parameters} \\[0.5em]
  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
  & \app && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra\Tm\,(\Gamma\ext \El\,a)\,B \\
  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma^\uparrow]) \\
  & {\app[]} && : (\app\,t)[\sigma^\uparrow] = \app\,(t[\sigma]) \\[0.5em]
  & \rlap{(4) Metatheoretic parameters} \\[0.5em]
  & \Pim && : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma \\
  & \blank\appm \blank && : \Tm\,\Gamma\,(\Pim\,T\,B)\ra(\alpha:T) \ra\Tm\,\Gamma\,(B\,\alpha) \\
  & {\Pim[]} && : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma]) \\
  & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha \\[0.5em]
\end{alignat*}
\subsubsection{Notation and abbreviations}

@@

  %% & \Ty && : \Con\ra\Set \\
  %% & \mathsf{empty} && : \Con \\
  %% & \mathsf{ext} && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  %% & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Con \\
  %% & \mathsf{El} && : (\Gamma:\Con)\ra\Ty\,(\mathsf{ext}\,\Gamma\,(\mathsf{U}\,\Gamma))

@@

\todo{Write this section}

Show that the theory of IIT signatures as described in Section
\ref{sec:theory_of_signatures} exists, that is, it can be defined only
using indexed inductive types and a $\Prop$ sort.

Ambroise has done this in Agda with funext and UIP using an
extrinsic--intrinsic translation.

Ambroise formalisation: \url{https://github.com/amblafont/UniversalII/tree/cwf-syntax/Cwf}.

In this formalization, there is a discrepancy for the typing rule of application
compared to the one of Section~\ref{sec:theory_of_signatures}: it is not
\[
   \app  : \Tm\,\Gamma\,(\Pi\,a\,B)\ra\Tm\,(\Gamma\ext \El\,a)\,B
\]
but
\[
   \app  : \Tm\,\Gamma\,(\Pi\,a\,B)\ra(u : \Tm\,\Gamma\,(\El\,a))\ra\,\Tm\,\Gamma\,B[<u>].
\]
Actually, I started with the first version, but at some point I
changed for the second one because it made some stuff easier (although
I don't remember exactly what it was).

I think this should be doable with only UIP (or a $\Prop$-valued
equality type?) and not funext. This would be important for a
computational justification of IITs.

Unfortunately, it seems that funext is needed (see the use of funext in
Syntax.agda) for the metatheoretic/infinitary
parameters part (4-5), for example in the case of the metatheoretic
$ \Pim  : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma $. This needs to be
investigated more carefully.

\paragraph*{Construction}
\input{universaliit}

>>>>

Inductive-inductive types (IITs \cite{forsberg-phd}) allow the mutual
definition of a type and a family of types over the first one. IITs
were originally introduced to represent the well-typed syntax of type
theory itself, and the main example is still the Chapman-style
\cite{chapman09eatitself} syntax for a type theory. A minimal use case
is the IIT of contexts and types given by the following constructors.
\begin{alignat*}{5}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\Set \\
  & \mathsf{empty} && : \Con \\
  & \mathsf{ext} && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \mathsf{U} && : (\Gamma:\Con)\ra\Ty\,\Gamma \\
  & \mathsf{El} && : (\Gamma:\Con)\ra\Ty\,(\mathsf{ext}\,\Gamma\,(\mathsf{U}\,\Gamma))
%  & \mathsf{Pi} && : (\Gamma:\Con)(A:\Ty\,\Gamma)\ra\Ty\,(\mathsf{ext}\,\Gamma\,A)\ra\,\Ty\,\Gamma
\end{alignat*}
This type has two sorts, $\Con$ and $\Ty$. The $\mathsf{ext}$ constructor of
$\Con$ refers to $\Ty$ and the $\Ty$-constructor $\mathsf{U}$ refers to $\Con$,
hence the two sorts have to be defined at the same time. Moreover, $\Ty$ is
indexed over $\Con$. This precludes a reduction analogous to the reduction of
$\mathsf{isEven}$--$\mathsf{isOdd}$, as we would get a type indexed over
itself. Another unique feature of IITs (which also shows up in higher inductive
types \cite{HoTTbook}) is that later constructors can refer to previous
constructors: in our case, $\mathsf{El}$ mentions
$\mathsf{ext}$.
% \[
%   \mathsf{Con?} : (b:\mathsf{Bool})\ra\mathsf{if}\,b\,\mathsf{then}\,\Set\,\mathsf{else}\,(\mathsf{Con?}\,\mathsf{false}\ra\Set),
% \]

@@

Reducing IITs to inductive types is an open problem. Forsberg
\cite{forsberg-phd} showed a reduction which only supports a simpler,
non-recursive-recursive elimination principle. Hugunin \cite{jasper} reduced
several IITs to inductive types, working inside a cubical type theory, but he
also only constructed a weaker eliminator.

The theory of signatures in \cite{Kaposi:2019:CQI:3302515.3290315} is a small
type theory in which every context encodes a valid quotient inductive-inductive
type (QIIT) signature. For example, the signature for natural numbers is simply
the context $(Nat:\U,zero:Nat,suc:Nat\ra Nat)$ of length three ($Nat$, $zero$
and $suc$ are variable names). \cite{Kaposi:2019:CQI:3302515.3290315} showed
that if this theory of signatures exists, then any finitary QIIT can be derived
from it via a term model construction. If we omit the type former for equality
constructors from the theory of signatures, we get a theory of finitary IIT
signatures, and the construction of \cite{Kaposi:2019:CQI:3302515.3290315} still
works: constructing an arbitrary IIT from the theory of IIT signatures does not
need the $\mathsf{Id}$ type former to be present.

However, the QIIT term model construction assumes that we already have the
syntax of QIIT signatures, and we can do induction over it. Hence, it only shows
that all finitary QIITs are derivable from a particular fairly complicated QIIT.
Likewise, restriced to finitary IITs, we only know that they are constructible
from a particular QIIT.

In this paper, we construct the theory of IIT signatures, which is a large
infinitary QIIT, using only indexed inductive types, uniqueness of identity
types (UIP) and function extensionality. Thus, we show that all finitary IITs are
reducible to the same feature set.

We first define a theory of IIT signatures in which every context is a signature
(Section \ref{sec:theory_of_signatures}). We present this as an algebraic
\cite{ttintt} theory, specifically as a quotient inductive-inductive type
(QIIT). In Section \ref{sec:ambroise} we show that this QIIT can be constructed
using inductive types, UIP and function extensionality. In Section
\ref{sec:constructingiits} we describe the construction of IITs from the syntax
of IIT signatures.

%% To derive the existence of all IITs from the theory of IIT signatures
%% we use a variant of the CwF model
%% \cite{Kaposi:2019:CQI:3302515.3290315} where types are modelled not by
%% displayed CwFs but by CwF isofibrations. We need this extension to
%% support infinitary constructors. Furthermore, using the method
%% introduced by Streicher \cite{streichersemantics} we show the
%% existence of the theory of IIT signatures. As opposed to Streicher, we
%% don't use quotients in this construction as the theory of IIT
%% signatures can be given only using neutral terms.

%% Our metatheory is ETT. Using Hofmann \cite{}, Theo \cite{} translation
%% we can translate our arguments to ITT+UIP+funext. Section
%% \ref{sec:ambroise} was formalised in Agda using UIP, rewrite rules and
%% funext.

%% This is the paper version of our TYPES 2019 talk.

@@

@@

The theory of IIT signatures is given as a category with families
(CwF)\cite{Dybjer96internaltype} with certain additional type formers. The CwF
part constitutes the substitution calculus, while the type formers allow
declaring new sorts and operators in a signature. Below, we list all components
for the theory of signatures; this can be viewed as a notion of model for the
theory, which is an iterated $\Sigma$ type. In this paper, we construct the
\emph{initial} such model, i.e.\ a model equipped with a unique recursion principle.

Note that initiality refers to a notion of homomorphism between models. As
of now, there is no general metatheory for (large, infinitary) QIITs from which
this notion would fall out. Hence, we give an ad-hoc definition for
homomorphisms, which can be summarized as a functor between underlying
categories which strictly preserves all additional family structure and type
formers.

\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \Con && : \Set \\
  & \Ty  && : \Con\ra\Set \\
  & \Sub  && : \Con\ra\Con\ra\Set \\
  & \Tm  && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set \\
  & \id && : \Sub\,\Gamma\,\Gamma \\
  & \blank\circ\blank && : \Sub\,\Theta\,\Delta\ra\Sub\,\Gamma\,\Theta\ra\Sub\,\Gamma\,\Delta \\
  & \ass && : (\sigma \circ \delta) \circ \nu = \sigma \circ (\delta \circ \nu) \\
  & \idl && : \id\circ\sigma = \sigma \\
  & \idr && : \sigma\circ\id = \sigma \\
  & \blank[\blank] && : \Ty\,\Delta\ra\Sub\,\Gamma\,\Delta\ra\Ty\,\Gamma \\
  & \blank[\blank] && : \Tm\,\Delta\,A\ra(\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma]) \\
  & [\id] && : A[\id] = A \\
  & [\circ] && : A[\sigma \circ \delta] = A[\sigma][\delta] \\
  & [\id] && : t[\id] = t \\
  & [\circ] && : t[\sigma \circ \delta] = t[\sigma][\delta] \\
  & \boldsymbol{\cdot} && : \Con \\
  & \epsilon && : \Sub\,\Gamma\,\boldsymbol{\cdot} \\
  & {\boldsymbol{\cdot}\eta} && : \{\sigma : \Sub\,\Gamma\,\boldsymbol{\cdot}\} \ra \sigma = \epsilon \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \blank,\blank && : (\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma])\ra\Sub\,\Gamma\,(\Delta\ext A) \\
  & \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
  & \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
  & {\ext\beta_1} && : \p\circ(\sigma, t) = \sigma \\
  & {\ext\beta_2} && : \q[\sigma, t] = t \\
  & {\ext\eta} && : (\p, \q) = \id \\
  & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\[0.5em]
  & \rlap{(2) Universe} \\[0.5em]
  & \U && : \Ty\,\Gamma \\
  & \El && : \Tm\,\Gamma\,\U\ra\Ty\,\Gamma \\
  & {\U[]} && : \U[\sigma] = \U \\
  & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\[0.5em]
  & \rlap{(3) Inductive parameters} \\[0.5em]
  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
  & \app && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra\Tm\,(\Gamma\ext \El\,a)\,B \\
  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma^\uparrow]) \\
  & {\app[]} && : (\app\,t)[\sigma^\uparrow] = \app\,(t[\sigma]) \\[0.5em]
  & \rlap{(4) Metatheoretic parameters} \\[0.5em]
  & \Pim && : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma \\
  & \blank\appm \blank && : \Tm\,\Gamma\,(\Pim\,T\,B)\ra(\alpha:T) \ra\Tm\,\Gamma\,(B\,\alpha) \\
  & {\Pim[]} && : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma]) \\
  & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha \\[0.5em]
\end{alignat*}
\subsubsection{Notation and abbreviations}

@@

@@

%% \todo{Write this section}

%% Show that the theory of IIT signatures as described in Section
%% \ref{sec:theory_of_signatures} exists, that is, it can be defined only
%% using indexed inductive types and a $\Prop$ sort.

%% Ambroise has done this in Agda with funext and UIP using an
%% extrinsic--intrinsic translation.

%% Ambroise formalisation: \url{https://github.com/amblafont/UniversalII/tree/cwf-syntax/Cwf}.

%% In this formalization, there is a discrepancy for the typing rule of application
%% compared to the one of Section~\ref{sec:theory_of_signatures}: it is not
%% \[
%%    \app  : \Tm\,\Gamma\,(\Pi\,a\,B)\ra\Tm\,(\Gamma\ext \El\,a)\,B
%% \]
%% but
%% \[
%%    \app  : \Tm\,\Gamma\,(\Pi\,a\,B)\ra(u : \Tm\,\Gamma\,(\El\,a))\ra\,\Tm\,\Gamma\,B[<u>].
%% \]
%% Actually, I started with the first version, but at some point I
%% changed for the second one because it made some stuff easier (although
%% I don't remember exactly what it was).

%% I think this should be doable with only UIP (or a $\Prop$-valued
%% equality type?) and not funext. This would be important for a
%% computational justification of IITs.

%% Unfortunately, it seems that funext is needed (see the use of funext in
%% Syntax.agda) for the metatheoretic/infinitary
%% parameters part (4-5), for example in the case of the metatheoretic
%% $ \Pim  : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma $. This needs to be
%% investigated more carefully.

\input{universaliit}

### 2019-11-11 paper/universaliit.tex
stuff

The construction consists of the following steps:
\begin{enumerate}
\item Definition of untyped syntax (as a family of inductive datatypes) together
  with typing judgments (as inductive relations on the untyped
  syntax), and construction of a model of the theory of IIT
  signatures from well-formed terms, denoted $S$.
\item Construction of a morphism $\rec:S\to M$ for arbitrary $M$, by:
  \begin{enumerate}
  \item defining a relation $\blank \sim \blank$ between the (well-typed) syntax and a given
    model. The idea is that given a syntactic context $\Gamma$ and a semantic
    context $\model{\Gamma}$ of the model $M$, we have $\Gamma \sim
    \model{\Gamma}$ if and only if $\rec\,\,\Gamma = \model{\Gamma}$, and
    similarly for types, terms, and substitutions;
    % semantic context $\model{\Gamma}$ of
    % the model $M$ if and only if $\Gamma$ relates to $\rec\,{\Gamma}$;
    % enjoyed by the initial morphism from the
    % syntax to a given model;
  \item showing that this relation is functional.
    \end{enumerate}
  \item Proving the uniqueness of this morphism by showing that any morphism
    $f:S\to M$ satisfies the relation. For example, for any syntactic context
    $\Gamma$ we have $\Gamma\sim f\,\Gamma$.
      % TODO donner composantes sur les contextes
\end{enumerate}
The next sections detail each of these steps.
\subsection{Syntactic Model}

@@

None of the constructions in this section use UIP. Functional extensionality is
necessary because the untyped metatheoretic $\Pi$ takes a metatheoretic function
as an argument. An example induction step that uses it is in the type
preservation proof for identity substitutions (this is an equation satisfied by
a model of the QIIT), in particular in the case $(\Pim\,T\,
A)[\id]=\Pim\,T\,A$. Indeed, the left hand side of this equation is equal to
$\Pim\,T\,(\lambda t.(A\,t)[\id])$ by definition, whereas the induction
hypothesis states that $(t:T)\ra (A\,t)[\id]=A\,t$.

@@

This approach is an alternative version of Streicher's method for interpreting
preterms in an arbitrary model\cite{streichersemantics}. Streicher first defines
a family of partial maps from the presyntax to a model, then shows that the maps
are total on well-formed input. We have found that our approach is significantly
easier to formalise. Too see why, note that the right notion of partial map in
type theory, which does not presume decidable definedness, is fairly
heavyweight:
\[
  \mathsf{PartialMap}\,A\,B := A\ra (P : \mathsf{hProp})\times(P \ra B)
\]
Here $\mathsf{hProp}$ denotes a type which is propositional. In the above
definition, we notice an opportunity for converting a fibered definition of a
type family into an indexed one; if we drop the propositionality for $P$ for the
time being, we may equivalently return a family indexed over $B$, which is
exactly just a relation $A\ra B\ra\Set$. Then, in our approach, we recover
uniqueness of $P$ through the functionality requirement on the $A\ra B\ra\Set$
relation, and totality by already assuming well-formedness of $A$. In type
theory, using indexed families instead of display maps is a common convenience,
since the former are natively supported in type theory, while the latter require
carrying around auxiliary propositional equalities.

@@

  & \rlap{(3) Parameters} \\[0.5em]

@@

Next, we prove that this relation is right unique.
Then, we show that the relation is stable under weakening and substitution.
  The last step consists of giving a related semantic counterpart to any
  well-typed context, type or term.
  Everything is done by induction on the typing judgments.

@@

  We show by recursion that the relation is right unique in the following sense:

@@

\subsection{The Recursor}
For the recursor, we build a morphism from the syntactic model to the semantic
one. The image of a syntactic context is a unique semantic context which is
related to it, and similarly for types and terms.
% We construct it as the semantic
% Induction on the typing judgments shows that any QIIT morphism from the
% syntactic model $S$  to a semantic model $M$ sends well-typed syntax on a related
% semantic counter-part.
Thus, as a first step, we use induction on well-formedness judgments to construct
semantic counterparts:
  \begin{alignat*}{5}
     & \Sigma\Con{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    \sum_{\model{\Gamma}} \wf{\Gamma} \sim \model{\Gamma}
    \\
     & \Sigma\Ty{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ({\model{A}:\model{\Ty}\model\Gamma})\times ( \wf{A} \sim \model{A})
    \\
     & \Sigma\Tm{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{t} : \Gamma \vdash t \in A ) \ra
    ({\model{t}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{t} \sim \model{t})
    \\
     & \Sigma\Var{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{x} : \Gamma \vdash x \invar A ) \ra
    ({\model{x}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{x} \sim \model{x})
    \\
     & \Sigma\Sub{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{\Delta} : \Delta \vdash  ) \ra
    ( \wf{\Delta} \sim \model{\Delta}) \ra
    \\ & && \qquad
    (\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta ) \ra
    ({\model{\sigma}:\model{\Sub}\model\Gamma}\model{\Delta})\times ( \wf{\sigma} \sim \model{\sigma})
  \end{alignat*}
  The right uniqueness of the relation is used in this induction. It is then
straightforward to show (without induction) that the first projection of these
constructions yield a model morphism from the syntax to the model, again using
right uniqueness.
\subsection{Uniqueness}
Our goal is to show that the syntactic model is initial. Thus, it remains to
show that the morphism constructed in the previous section is unique. We exploit
right uniqueness of the relation: it is enough to show that any such morphism
maps a syntactic context to a related semantic context, and similarly for types
and terms.

>>>>

The construction consists of the following steps:

\begin{enumerate}
\item Definition of untyped syntax (as a family of inductive datatypes) together
  with typing judgments (as inductive relations on the untyped syntax), and
  construction of a model of the theory of IIT signatures from well-formed
  terms, denoted $S$.
\item Construction of a morphism $\rec:S\to M$ for arbitrary $M$, by:
  \begin{enumerate}
  \item defining a relation $\blank \sim \blank$ between the well-typed syntax
    and a given model. The idea is that given a syntactic context $\Gamma$ and a
    semantic context $\model{\Gamma}$ of the model $M$, we have
    $\Gamma \sim \model{\Gamma}$ if and only if $\rec\,\,\Gamma
    = \model{\Gamma}$, and similarly for types, terms, and substitutions;
    % semantic context $\model{\Gamma}$ of
    % the model $M$ if and only if $\Gamma$ relates to $\rec\,{\Gamma}$;
    % enjoyed by the initial morphism from the
    % syntax to a given model;
  \item showing that this relation is functional.
    \end{enumerate}
  \item Proving the uniqueness of this morphism by showing that any morphism
    $f:S\to M$ satisfies the relation. For example, for any syntactic context
    $\Gamma$ we have $\Gamma\sim f\,\Gamma$.
      % TODO donner composantes sur les contextes
\end{enumerate}
The next sections detail each of these steps.
\subsection{Syntactic Model}

@@

Up until the construction of the syntactic model, UIP is not used. Function
extensionality on the other hand is necessary because the untyped metatheoretic
$\Pi$ takes a metatheoretic function as an argument. An example induction step
that uses it is in the type preservation proof for identity substitutions (this
is an equation satisfied by a model of the QIIT), in particular in the case
$(\Pim\,T\, A)[\id]=\Pim\,T\,A$. Indeed, the left hand side of this equation is
equal to $\Pim\,T\,(\lambda t.(A\,t)[\id])$ by definition, whereas the induction
hypothesis states that $(t:T)\ra (A\,t)[\id]=A\,t$.

@@

This approach is an alternative version of Streicher's method for interpreting
preterms in an arbitrary model\cite{streichersemantics}. Streicher first defines
a family of partial maps from the presyntax to a model, then shows that the maps
are total on well-formed input. We have found that our approach is significantly
easier to formalise. Too see why, note that the right notion of partial map in
type theory, which does not presume decidable definedness, is fairly
heavyweight:
\[
  \mathsf{PartialMap}\,A\,B := A\ra (P : \Set)\times\isaprop\,P\times (P \ra B)
\]
In the above definition, we notice an opportunity for converting a fibered
definition of a type family into an indexed one; if we drop the propositionality
for $P$ for the time being, we may equivalently return a family indexed over
$B$, which is exactly just a relation $A\ra B\ra\Set$. Then, in our approach, we
recover uniqueness of $P$ through the functionality requirement on the $A\ra
B\ra\Set$ relation, and totality by already assuming well-formedness of $A$. In
type theory, using indexed families instead of display maps is a common
convenience, since the former are natively supported in type theory, while the
latter require carrying around auxiliary propositional equalities.

@@

  & \rlap{(3) Inductive parameters} \\[0.5em]

@@

Next, we prove that this relation is right unique.  Then, we show that the
relation is stable under weakening and substitution.  The last step consists of
showing left-totality, i.e.\ giving a related semantic counterpart to any
well-typed context, type or term.  Everything is done by induction on the typing
judgments.

@@

  We show by induction that the relation is right unique in the following sense:

@@

\subsection{Left-Totality and the Recursor}
For the recursor, we build a morphism from the syntactic model to the semantic
one. The image of a syntactic context is a unique semantic context which is
related to it, and similarly for types and terms.
% We construct it as the semantic
% Induction on the typing judgments shows that any QIIT morphism from the
% syntactic model $S$  to a semantic model $M$ sends well-typed syntax on a related
% semantic counter-part.
Thus, as a first step, we use induction on well-formedness judgments to construct
semantic counterparts:
  \begin{alignat*}{5}
     & \Sigma\Con{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    \sum_{\model{\Gamma}} \wf{\Gamma} \sim \model{\Gamma}
    \\
     & \Sigma\Ty{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ({\model{A}:\model{\Ty}\model\Gamma})\times ( \wf{A} \sim \model{A})
    \\
     & \Sigma\Tm{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{t} : \Gamma \vdash t \in A ) \ra
    ({\model{t}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{t} \sim \model{t})
    \\
     & \Sigma\Var{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{x} : \Gamma \vdash x \invar A ) \ra
    ({\model{x}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{x} \sim \model{x})
    \\
     & \Sigma\Sub{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{\Delta} : \Delta \vdash  ) \ra
    ( \wf{\Delta} \sim \model{\Delta}) \ra
    \\ & && \qquad
    (\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta ) \ra
    ({\model{\sigma}:\model{\Sub}\model\Gamma}\model{\Delta})\times ( \wf{\sigma} \sim \model{\sigma})
  \end{alignat*}
  The right uniqueness of the relation is used in this induction. It is then
straightforward to show (without induction, but using right uniqueness) that the
first projection of these constructions yield a model morphism from the syntax
to the model.

\subsection{Uniqueness}
Our goal is to show that the syntactic model is initial. Thus, it remains to
show that the morphism constructed in the previous section is unique. We exploit
right uniqueness of the relation: it is enough to show that any such morphism
maps a syntactic context to a related semantic context, and similarly for types
and terms.
